---
title: "Using `infer` package for Statistical Inference"
author: "Francisco Cruz (francruz)"
date: "1/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 3.5)
options(digits = 4)
```

## Introduction

## Data Preparation

I will use `nycflights13` to support several typical statistical tests. `dplyr` will also be used extensively to integrate the full workflow within the `tidyverse` grammar.

```{r loading libraries, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# Reference: https://infer-dev.netlify.com/articles/observed_stat_examples.html
library(nycflights13)
library(dplyr)
library(ggplot2)
library(stringr)
library(infer)
set.seed(2020)
```

First, we prepare a dataframe with the following structure, allowing us to perform the most typical statistical analyses:

- Two numeric: `arr_delay`, `dep_delay`
- Two categorical:
    - `season` (`winter`, `summer`)
    - `day_hour` (`morning`, `not morning`)
- Three categorical: `origin` (`EWR`, `JFK`, `LGA`)
- Sixteen categorical: `carrier`

```{r echo=TRUE, message=FALSE, warning=FALSE}
fli_small <- flights %>% 
    na.omit() %>% 
    sample_n(size = 500) %>% 
    mutate(season = case_when(month %in% c(10:12, 1:3) ~ 'winter',
                              month %in% c(4:9) ~ 'summer')) %>% 
    mutate(day_hour = case_when(between(hour, 1, 12) ~'morning',
                                between(hour, 13,24) ~'not morning')) %>%     select(arr_delay, dep_delay, season, day_hour, origin, carrier)
```

## Hypothesis Tests

### One Numerical Value (mean)

Let's look at the mean of the departure delay of flights in our dataset. We can clculate this value in different ways; here are just a few. Notice that the last one makes use of the `infer` package:

```{r calculate point average, echo=TRUE, message=TRUE, warning=FALSE}
x_bar <- mean(fli_small[['dep_delay']])
x_bar <- mean(fli_small$dep_delay)

(x_bar <- fli_small %>% 
        specify(response = dep_delay) %>% 
        calculate(stat = 'mean'))
```

Let's now generate the null distribution based on an initial hypothesis that the average departure delay is 10 hours. This would be a 'point' test.

```{r}
null_dist <- fli_small %>% 
    specify(response = dep_delay) %>% 
    hypothesize(null = 'point', mu = 10) %>% 
    generate(reps = 1000) %>% 
    calculate(stat = 'mean')

visualize(null_dist) +
    shade_p_value(obs_stat = x_bar,
                  direction = 'two_sided')

significance <- null_dist  %>%
    get_p_value(obs_stat = x_bar, direction = 'two_sided')
```

Notice that the grammar follows the inferential workflow: we `specify` the value over which we are running the test, and `hypothesize` our null $H_{0}: \mu = 10$. Next, we `generate` 1000 replicates of this null hypotesis and `calculate` the mean `dep_delay` for each one.

We use `visualize` as a wrapper for `ggplot` to visualize our test, and since our null is expressed as an equality, the test is `two-sided` to represent the alternate hypothesis $H_{a} \neq 0$. Passing the observed mean as `obs_stat = x_bar` sets the critical point for the test.

Finally, `get_p_value` outputs the significance of the test. In this case we obtain a value of `r significance` which is greater than 0.05, so we don't have evidence to reject our null hypothesis that the average departure delay is 10 minutes.

### One Numerical Variable (standardized mean $t$)

Reproducing a similar analysis, but this time applying the test over the standardized mean, and testing against the null hypothesis $H_{0}: \mu = 8$

```{r}
(t_bar <- fli_small %>% 
        t_stat(response = dep_delay, mu = 8))
```

Following the same workflow:

```{r}
null_dist <- fli_small %>% 
    specify(response = dep_delay) %>% 
    hypothesize(null = 'point', mu = 8) %>% 
    generate(reps = 1000) %>% 
    calculate(stat = 't')

visualize(null_dist) +
    shade_p_value(obs_stat = t_bar,
                  direction = 'two_sided')

significance <- null_dist  %>%
    get_p_value(obs_stat = t_bar, direction = 'two_sided')
```

In this case we obtain a p-value of `r significance` which is just barely greater than 0.05.

### One Numerical Variable (median)

As usual, we first calculate the statistic from our sample

```{r}
(x_tilde <- fli_small %>% 
     specify(response = dep_delay) %>% 
     calculate(stat = 'median') %>% 
     pull())
```

And now calculate the value of the stat across 1000 replicates to represent our null hypothesis, $H_{0}: \tilde{x} = -1$.

```{r}
null_dist <- fli_small %>% 
    specify(response = dep_delay) %>% 
    hypothesize(null = 'point', med = -1) %>% 
    generate(reps = 1000) %>% 
    calculate(stat = 'median')

visualize(null_dist) +
    shade_p_value(obs_stat = x_tilde, direction = 'two_sided')
```

And the p-value is calculated as:

```{r}
significance <- null_dist %>% 
    get_p_value(obs_stat = x_tilde, direction = 'two_sided') %>% 
    pull()
```

giving a value of `r significance`, which allows us to reject the null hypothesis and affirm that a median of -1 is unlikely given the sample distribution.

### One Categorical (a proportion)

Let's look now at how to test for a proportion. We will check whether there is no difference in the proportion of delayed departures between morning and non-morning flights. In our sample, this proportion is:

```{r}
(p_hat <- fli_small %>% 
     specify(response = day_hour, success = 'morning') %>% 
     calculate(stat = 'prop'))
```

So, we see that in our sample, the proportion of delayed morning flights is `r p_hat * 100` %. Is this an indication of more delays in non-morning flights? Well, our null hypothesis would be that there is no difference in this proportion, i.e.: $H_{0}: \hat{p}=0.5$. Let's build 1000 replicates of such case and test agains our observation

```{r}
null_dist <- fli_small %>% 
    specify(response = day_hour, success = 'morning') %>% 
    hypothesize(null = 'point', p = 0.5) %>% 
    generate(reps = 1000) %>% 
    calculate(stat = 'prop')

visualize(null_dist) +
    shade_p_value(obs_stat = p_hat, direction = 'two_sided')
```

From the visualization above, we can guess that it is unlikely that the delays happen equally in morning and non-morning flights based on the observed proportion in our sample. To calculate the p-value of this proportion:

```{r}
significance <- null_dist %>% 
    get_p_value(obs_stat = p_hat, direction = 'two_sided') %>% 
    pull()
```

And at `r significance` -- as expected -- it is significantly lower than 0.05. The null hypothesis can be rejected.

### Two categorical variables (2 levels)

We will look now at a different type of test: in our sample we have now a categorical variable `season` with two levels ('winter' and 'summer'). We'd like to see if there is a difference in proportion of morning flights between those two seasons. Our response variable `day_hour` is going to be evaluated for those two levels of `seasson`.

Since we have two different levels, we calculate the proportions of morning flights for each of the two seasons, and look at the difference between them. We will test whether the `day_hour` is independent from the level of `season` in our sample:

```{r}
(d_hat <- fli_small %>% 
     specify(day_hour ~ season, success = 'morning') %>% 
     calculate(stat = 'diff in props', order = c('winter', 'summer')))
```

Pay notice to the formula in `specify`. Since we are considering the proportions of each value of `day_hour` for each level of `season` the relation is given as the formula `day_hour ~ season`). The `order` parameter is used to indicate how the difference is calculated (`winter` - `summer`).

The null distribution is built as

```{r}
null_dist <- fli_small %>% 
    specify(day_hour ~ season, success = 'morning') %>% 
    hypothesize(null = 'independence') %>% 
    generate(reps = 1000) %>% 
    calculate(stat = 'diff in props', order = c('winter', 'summer'))

visualise(null_dist) +
    shade_p_value(obs_stat = d_hat, direction = 'two_sided')

significance <- null_dist %>% 
    get_p_value(obs_stat = d_hat, direction = 'two_sided') %>% 
    pull()
```

Both from the graph and from the p-value (`r significance`) we see that the null hypothesis can't be rejected: the observed difference is very likely to come from a distribution where `day_hour` doesn't depend on `season`.

